Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-06 exp1/iswc2013.tex
comment pending tripleC disaster
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@515 43ef1190-90fd-4dd7-a769-d0121f02b80e

To proceed beyond a visual judgement of significance, we may use
Lenth's method\todo{ref Lenth}, which is a simple method for
estimating significance in an experiment with no replicates. We plot
the results in figure~\ref{fig:fulllenth}. Visually, this is not as
useful, since the large number\todo{really include?} We see that we
have very significant effects. We also note that ``Implement'' itself
is significantly negative. We also note that the ``Implement:Lang''
interaction is significantly positive, as expected from our setup,
where ``Lang'' was the detrimental side-effect we simulated.

To investigate whether the detrimental side-effects cancels the
positive effect of our simulated optimization, we have to formulate a
hypothesis test. By inspecting the normal plot in figure~\ref{fig:fullnormal}

>>>>

To proceed beyond a visual judgement of significance, we may use
Lenth's method\todo{ref Lenth}, which is a simple method for
estimating significance in an experiment with no replicates. We plot
the results in figure~\ref{fig:fulllenth}. In
figure~\ref{fig:fullnormal}, Lenth's method is used to label the
significant effects. Visually, this is not as useful, since the large
number\todo{really include?} We see that we have very significant
effects. We also note that ``Implement'' itself is significantly
negative. We also note that the ``Implement:Lang'' interaction is
significantly positive, as expected from our setup, where ``Lang'' was
the detrimental side-effect we simulated.

To investigate whether the detrimental side-effects cancels the
positive effect of our simulated optimization, we have to formulate a
hypothesis test. By inspecting the normal plot in
figure~\ref{fig:fullnormal}, we see that only five factors are highly
significant by either being a significant main effect or participate
in a highly significant interaction. We say that the three factors
``BGPComp'', ``Range'' and ``Machine'' are \emph{inactive} since they
do not contribute significantly to the observed variation in
performance. In practical situations, the four factors ``Lang'',
``Optional'', ``Union'' and ``TripleC'' can be regarded as given,
i.e. we cannot change the size of the dataset without loss of
information, or change the query, since it gives a different
answer. In practice, we can only control ``Implement'', i.e. we can
only change the implementation. In this context, we call ``Implement''
a \emph{control} factor and the other four active factors
\emph{environmental}.

\todo{finish the t-test stuff once ready}

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are critical. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. This should cause some alarm. It is a fundamental
property of two-level experiments that it can only model linear
effects. If the effect is truly linear, and the variance is small, the
two levels can be quite close together and still yield a reasonable
estimate for the effect. However, if the effect is non-linear, the
choice of levels becomes critical. In this case, the experimenter
should be concerned that they may have a non-linear effect that is not
well represented by the sizes of datasets. Such reasoning should be
applied to all significant effects.
